---
title: " "
format: 
  pdf:
    include-in-header: 
      text: |
        \usepackage{setspace}
        \usepackage{typearea}
    title-meta: 'More Than Just Associations: An Introduction to Causal Inference for Sport Science'
    author-meta: 'Simon Nolte'
    subject: 'Released under a CC-BY-NC-ND license'
    linestretch: 1.5
number-sections: true
top-level-division: section
mainfont: Arial
fontsize: 11pt
urlcolor: blue
linkcolor: black
link-citations: true
bibliography: ../references.bib
csl: '../apa.csl'
geometry: "left = 4cm, right = 2cm, top = 2cm, bottom = 2cm"
execute: 
  echo: false
params:
  title: 'More Than Just Associations: An Introduction to Causal Inference for Sport Science'
  name: Simon Nolte
  university: German Sport University Cologne
  year: 2024
---

\singlespacing

\begin{centering}

\vspace*{2cm}

\LARGE
{`r params$title`}

\vspace{5cm}

\LARGE
{Master thesis}

\LARGE
{From}

\vspace{2cm}

\LARGE
{`r params$name`}

\vspace{6cm}

\LARGE
{`r params$university`}

\Large
{`r paste0("Cologne ", params$year)`}

\end{centering}

\pagenumbering{gobble}

\newpage

Thesis supervisor:

Dr. Oliver Jan Quittmann

Institute of Movement and Neurosciences

\newpage

Affirmation in lieu of an oath

Herewith I affirm in lieu of an oath that I have authored this Bachelor thesis independently and did not use any other sources and tools than indicated. All citations, either direct quotations or passages which were reproduced verbatim or nearby-verbatim from publications, are indicated and the respective references are named. The same is true for tables and figures. I did not submit this piece of work in the same or similar way or in extracts in another assignment.


\_____________________________

Personally signed

\newpage

\onehalfspacing

```{r}
#| child = 'abstract.qmd'
```

\newpage

\singlespacing

\pagenumbering{roman}

# Table of Contents {-}

\renewcommand{\contentsname}{}

\tableofcontents

\newpage

\renewcommand{\listfigurename}{}

# List of Figures {-}

\listoffigures

\vspace{2cm}

\renewcommand{\listtablename}{}

# List of Tables {-}

\listoftables

\newpage

\pagenumbering{arabic}

\onehalfspacing

```{r}
#| child = 'introduction.qmd'
```

\newpage

```{r}
#| child = 'theory.qmd'
```

\newpage

```{r}
#| child = 'methods.qmd'
```

\newpage

```{r}
#| child = 'results.qmd'
```

\newpage

```{r}
#| child = 'discussion.qmd'
```

\newpage

```{r}
#| child = 'conclusion.qmd'
```

\newpage

# References {.unnumbered}

<div id="refs"></div>

\newpage

\appendix

# Appendix

## Mathematical Background

```{r}
#| child = 'math.qmd'
```

\newpage

## Simulations  {#sec-sim}

```{r}
#| child = 'simulations.qmd'
```


\newpage

## Technical Details {#sec-tech}

### Session Info

```{r session}
#| echo: true
sessionInfo()
```

### Packages

```{r}
#| message: false
#| echo: true

# p_used <- suppressMessages(unique(renv::dependencies(path = "../")$Package))
# p_inst <- as.data.frame(installed.packages())
# out <- p_inst[p_inst$Package %in% p_used, c("Package", "Version")]
# rownames(out) <- NULL
# out
```

